\documentclass{article}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}

\title{\textbf{CSE400 – Fundamentals of Probability in Computing}}
\author{\textbf{Kensi Patel - AU2440102}}
\date{Lecture 23}

\begin{document}

\maketitle

\textbf{Lecture 23: Predicting the Impossible – Tail Bounds and System Reliability}

Instructor: Dhaval Patel, PhD

\section*{1. Problem Context (Amazon Black Friday Case Study)}

\begin{itemize}
    \item \textbf{System:} Large-scale distributed servers (Amazon sale traffic).
    \item \textbf{Model:} Requests arrive randomly over time $\rightarrow$ modeled using \textbf{Poisson distribution}. (Page 3)
\end{itemize}

\textbf{Let:}
\begin{itemize}
    \item $X$ = number of requests in a time interval
    \item $\lambda$ = average rate (mean)
\end{itemize}

\textbf{From slide:}
\begin{itemize}
    \item Example: $\lambda = 50$ requests/min (Page 5)
    \item Crash threshold: $k = 60$
\end{itemize}

\textbf{Objective}

Compute:
\[P(X \ge k)\]

$\rightarrow$ Probability system overloads.

\section*{2. Key Observation from Lecture}

From \textit{Page 5}:
\begin{itemize}
    \item ``The mean is safe. The tail is catastrophic.''
\end{itemize}

\textbf{Meaning:}
\begin{itemize}
    \item Average load does not cause failure.
    \item Rare high spikes (tail events) cause crashes.
\end{itemize}

\textbf{Thus:}
\begin{itemize}
    \item We must estimate tail probability $P(X \ge k)$.
\end{itemize}

\section*{3. Need for Bounds}

Exact probability is difficult or expensive to compute.

From \textit{Page 6}:
\begin{itemize}
    \item Use upper bounds instead of exact values.
    \item Better information $\rightarrow$ tighter bounds.
\end{itemize}

\textbf{Hierarchy:}
\begin{enumerate}
    \item Markov $\rightarrow$ uses only mean
    \item Chebyshev $\rightarrow$ uses mean + variance
    \item Chernoff $\rightarrow$ uses full distribution (MGF)
\end{enumerate}

\section*{4. Markov's Inequality}

\textbf{Statement}

For non-negative random variable $X$:
\[P(X \ge a) \le \frac{E[X]}{a}\]

\textbf{Derivation (as implied in lecture)}

Using indicator:
\[X \ge a \cdot \mathbf{1}_{X \ge a}\]

Take expectation:
\[E[X] \ge a \cdot P(X \ge a)\]

Rearrange:
\[P(X \ge a) \le \frac{E[X]}{a}\]

\textbf{Properties (Page 7)}
\begin{itemize}
    \item Requires: Mean only
    \item Decay: $\frac{1}{k}$ (linear)
    \item Very loose (overestimates heavily)
\end{itemize}

\textbf{Example from slide:}
\begin{itemize}
    \item Exact $\approx 0.004$
    \item Markov $\approx 0.714$ (165$\times$ too high)
\end{itemize}
\section*{5. Chebyshev's Inequality}

\textbf{Statement}

\[P(|X - \mu| \ge t) \le \frac{\sigma^2}{t^2}\]

For right tail:
\[P(X \ge a) \le \frac{\sigma^2}{(a - \mu)^2}\]

\textbf{Derivation (based on Markov)}

Apply Markov to:
\[Y = (X - \mu)^2\]

Then:
\[P(|X - \mu| \ge t) = P(Y \ge t^2)\]

Using Markov:
\[P(Y \ge t^2) \le \frac{E[Y]}{t^2}\]

Since:
\[E[(X - \mu)^2] = \sigma^2\]

Final:
\[P(|X - \mu| \ge t) \le \frac{\sigma^2}{t^2}\]

\textbf{Properties (Page 8)}

Requires: Mean + Variance

Decay:$\frac{1}{k^2} \quad \text{(polynomial)}
$

Better than Markov

\textbf{Example:}

Exact $\approx 0.004$

Chebyshev $\approx 0.125$ (29$\times$ too high)

\section*{6. Chernoff Bound}

\textbf{Key Idea (Page 9)}

Use Moment Generating Function (MGF):
\[M_X(t) = E[e^{tX}]\]

\textbf{General Bound}
\[P(X \ge a) \le \min_{t > 0} e^{-ta} M_X(t)\]

\section*{Derivation Steps (Page 18 Blueprint)}
\begin{itemize}
    \item Step 1: Define exponential transform
    \[P(X \ge a) = P(e^{tX} \ge e^{ta})\]
    \item Step 2: Apply Markov
    \[P(e^{tX} \ge e^{ta}) \le \frac{E[e^{tX}]}{e^{ta}}= e^{-ta} M_X(t)\]
    \item Step 3: Optimize over $t$
    \[\min_{t > 0} e^{-ta} M_X(t)\]
\end{itemize}

\section*{Poisson Case (Page 9)}

\begin{itemize}
    \item For $X \sim \text{Poisson}(\lambda)$:
    \[M_X(t) = e^{\lambda (e^t - 1)}\]
    \item Final bound form:
    \[P(X \ge a) \le e^{-\lambda (r \ln r - r + 1)}, \quad r = \frac{a}{\lambda}\]
\end{itemize}

\section*{Properties}
\begin{itemize}
    \item Requires: Full distribution (MGF)
    \item Decay: Exponential $e^{-k}$
    \item Tightest bound
    
    \textbf{Example:}
    \item Exact $\approx 0.092$
    \item Chernoff $\approx 0.391$ (closest)
\end{itemize}

\section*{7. Comparison of Bounds (Page 10)}

\begin{tabular}{lll}
\toprule
\textbf{Bound} & \textbf{Decay} & \textbf{Tightness} \\
\midrule
Markov & $1/k$ & Very loose \\
Chebyshev & $1/k^2$ & Moderate \\
Chernoff & $e^{-k}$ & Tight \\
\bottomrule
\end{tabular}

\section*{8. System Design Interpretation}

\textbf{SLA Scenario (Page 11)}
\begin{itemize}
    \item $\lambda = 500$
    \item Threshold $k = 600$
\end{itemize}

\textbf{Goal:}
\[P(X \ge 600)\]

\section*{Impact of Bound Choice (Page 12)}

\begin{tabular}{ll}
\toprule
\textbf{Method} & \textbf{Estimated Failure} \\
\midrule
Exact & $\sim 1.4\%$ \\
Chernoff & $\sim 8.8\%$ \\
Chebyshev & $\sim 20\%$ \\
Markov & $\sim 90.9\%$ \\
\bottomrule
\end{tabular}

\section*{Conclusion from Lecture}

\begin{itemize}
    \item Loose bounds $\rightarrow$ huge overestimation
    \item Leads to:
    \begin{itemize}
        \item Over-provisioning
        \item Financial waste
    \end{itemize}
\end{itemize}

\section*{9. Waste Multiplier (Page 14)}

\[
\text{Waste} = \frac{\text{Bound Estimate}}{\text{Exact Probability}}
\]

\textbf{Interpretation:}
\begin{itemize}
    \item Larger bound $\Rightarrow$ more unnecessary servers purchased
\end{itemize}

\section*{10. Distribution Sensitivity (Page 15)}

\begin{itemize}
    \item Chernoff works well for:
    \begin{itemize}
        \item Poisson
        \item Binomial
    \end{itemize}
    \item Not accurate for:
    \begin{itemize}
        \item Heavy-tailed (Gamma)
    \end{itemize}
    $\rightarrow$ In such cases:
    \begin{itemize}
        \item Use simulations or Chebyshev
    \end{itemize}
\end{itemize}

\section*{11. Decision Matrix (Page 16)}

\begin{tabular}{lll}
\toprule
\textbf{Method} & \textbf{Requires} & \textbf{Use Case} \\
\midrule
Markov & Mean & Quick rough bound \\
Chebyshev & Mean + Variance & Unknown distribution \\
Chernoff & Full MGF & Precise system sizing \\
\bottomrule
\end{tabular}

\section*{12. Class Exercise (Laplace Distribution)}

\textbf{Given:}

\[X \sim \text{Laplace}(\mu, b), \quad \sigma = b\sqrt{2}\]

\textbf{Steps (Page 18)}

\begin{enumerate}
    \item Compute MGF:
    \[M_X(t) = E[e^{tX}]\]
    \item Apply Chernoff:
    \[P(X \ge a) \le e^{-ta} M_X(t)\]
    \item Optimize:
    \[\frac{d}{dt} (\cdot) = 0\]
    \item Substitute optimal $t$
\end{enumerate}

\section*{Final Takeaways}

\begin{itemize}
    \item Tail probability is critical for system reliability.
    \item Exact computation often replaced by bounds.
    \item Tightness hierarchy:
    \[\text{Markov} < \text{Chebyshev} < \text{Chernoff}\]
    \item Wrong bound $\Rightarrow$ wrong engineering decisions.
\end{itemize}















\end{document}